---
title: "Relationship between Crime and Social Factors in 1850s British Counties"
author: "Your Name"
output:
  pdf_document:
    latex_engine: xelatex
fontsize: 12pt
geometry: margin=2.5cm
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{Your Name}
  - \fancyhead[R]{\thepage}
  - \fancyfoot[C]{}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=5)
```

# Introduction

The goal of this analysis is to investigate the relationship between crime rates and certain social factors in mid-19th century British counties. The dataset is sourced from John Clay's 1857 study, examining education, religious worship, and the prevalence of beer houses in relation to crime.

## Variables:
- **Criminals per 100k population**: Number of convicted criminals per 100,000 inhabitants.
- **Ale/Beer Houses per 100k**: Density of public drinking establishments.
- **Attendants at school per 10k**: Proxy for education level.
- **Attendants at public worship per 2000**: Proxy for religious observance.

The scientific question is: *How do education, religious attendance, and beer house prevalence relate to crime rates in 19th-century British counties?*

---

# Data Loading and Cleaning

```{r libraies}
library(GGally)
```


```{r load-data}
url <- "http://users.stat.ufl.edu/~winner/data/beerhall.dat"
beer_data <- read.fwf(file=url, widths=c(16,-1,19,-6,1,-4,4,-4,4,-4,4,-4,4),
                      col.names=c("County","RegionName","RegionCode",
                                  "Criminals","BeerHouses","School","Church"))
beer_data$RegionCode <- factor(beer_data$RegionCode)
summary(beer_data)
```

---

# Exploratory Data Analysis (EDA)

## Univariate Summaries

```{r summary-stats}
summary(beer_data[, c("Criminals", "BeerHouses", "School", "Church")])
```

## Bivariate Correlations

# Linear relationships 
```{r correlation-matrix}
cor_matrix <- cor(beer_data[, c("Criminals", "BeerHouses", "School", "Church")])
print(round(cor_matrix, 2))
```
linear relationships: 
           Criminals BeerHouses School Church
Criminals       1.00       0.46  -0.23   0.00
BeerHouses      0.46       1.00   0.14   0.15
School         -0.23       0.14   1.00   0.60
Church          0.00       0.15   0.60   1.00


# Non linear relationships
## Compute mutual information for each pair of variables
```{r correlation-matrix}
mi_matrix <- mutinformation(beer_data[, c("Criminals", "BeerHouses", "School", "Church")])

# Print the mutual information matrix
print(round(mi_matrix, 2))
```

non-linear relationships
           Criminals BeerHouses School Church
Criminals       3.55       3.52   3.05   3.48
BeerHouses      3.52       3.65   3.16   3.58
School          3.05       3.16   3.19   3.12
Church          3.48       3.58   3.12   3.62

## Distribution of Crime Rates

```{r plot-histogram}
library(ggplot2)
ggplot(beer_data, aes(x=Criminals)) + 
  geom_histogram(bins=10, fill="blue", alpha=0.5) + 
  labs(title="Distribution of Crime Rates", x="Criminals per 100k", y="Frequency")
```

## Scatterplots

```{r scatterplots}
ggplot(beer_data, aes(x=BeerHouses, y=Criminals)) +
  geom_point() +
  geom_smooth(method="lm", col="blue") +
  labs(title="Crime Rate vs. Beer Houses", x="Beer Houses per 100k", y="Criminals per 100k")
```
```{r scatterplots}

ggplot(beer_data, aes(x=School, y=Criminals)) +
  geom_point() +
  geom_smooth(method="lm", col="red") +
  labs(title="Crime Rate vs. School Attendance", x="School Attendance per 10k", y="Criminals per 100k")
```


```{r scatterplots}

ggplot(beer_data, aes(x=Church, y=Criminals)) +
  geom_point() +
  geom_smooth(method="lm", col="#00ff7b") +
  labs(title="Crime Rate vs. Church Attendance", x="Church Attendance per 10k", y="Criminals per 100k")
```
```{r scatterplots}
# Histogram for Crime Rate (Criminals per 100k)
ggplot(beer_data, aes(x = Criminals)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Crime Rate (Criminals per 100k)",
       x = "Criminals per 100k", y = "Frequency")
```
```{r scatterplots}
# Histogram for Beer Houses per 100k
ggplot(beer_data, aes(x = BeerHouses)) +
  geom_histogram(binwidth = 50, fill = "darkorange", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Beer Houses per 100k",
       x = "Beer Houses per 100k", y = "Frequency")
```
```{r scatterplots}
# Histogram for School Attendance per 10k
ggplot(beer_data, aes(x = School)) +
  geom_histogram(binwidth = 50, fill = "darkgreen", color = "black", alpha = 0.7) +
  labs(title = "Histogram of School Attendance per 10k",
       x = "School Attendance per 10k", y = "Frequency")
```
```{r scatterplots}
# Histogram for Church Attendance per 2000
ggplot(beer_data, aes(x = Church)) +
  geom_histogram(binwidth = 50, fill = "purple", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Church Attendance per 2000",
       x = "Church Attendance per 2000", y = "Frequency")
```

```{r scatterplots}
# Scatterplot: Crime Rate vs. Beer Houses
ggplot(beer_data, aes(x = BeerHouses, y = Criminals)) +
  geom_point(color = "darkorange", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(title = "Crime Rate vs. Beer Houses",
       x = "Beer Houses per 100k", y = "Criminals per 100k")
```

```{r scatterplots}
# Scatterplot: Crime Rate vs. School Attendance
ggplot(beer_data, aes(x = School, y = Criminals)) +
  geom_point(color = "darkgreen", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(title = "Crime Rate vs. School Attendance",
       x = "School Attendance per 10k", y = "Criminals per 100k")
```

```{r scatterplots}
# Scatterplot: Crime Rate vs. Church Attendance
ggplot(beer_data, aes(x = Church, y = Criminals)) +
  geom_point(color = "purple", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(title = "Crime Rate vs. Church Attendance",
       x = "Church Attendance per 2000", y = "Criminals per 100k")
```

# Better analysis of correlation and features 
```{r scatterplot_matrix}
# Scatterplot Matrix (Pair Plot) for all numerical variables
ggpairs(beer_data[, c("Criminals", "BeerHouses", "School", "Church")],
        upper = list(continuous = wrap("cor", size = 5)),
        lower = list(continuous = wrap("smooth", method = "lm", color = "blue")))
```

Schools and churches are highly correlated with 0.597***, This and the non-linear trend
explored later seems to make churches a redondant and non-significant feature. It seems 
to be reasonable to concetrate on Schools and Beerhouses to create a sound and simple model


# Relationship of criminals and beerhouses 
```{r scatterplots_beerhouses_criminals}
ggplot(beer_data, aes(x = BeerHouses, y = Criminals)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "loess", color = "red", span = 0.75) + 
  labs(title = "Crime Rate vs Beer Houses (LOESS Curve)", 
       x = "Beer Houses per 100k", y = "Criminals per 100k")

```
seems to have a logaritimic relationship

# log transfomration of the beerhouses 
```{r scatterplots_beerhouses_criminals}
ggplot(beer_data, aes(x = log(BeerHouses), y = Criminals)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "loess", color = "red", span = 0.75) + 
  labs(title = "Crime Rate vs Beer Houses (LOESS Curve)", 
       x = "Beer Houses per 100k", y = "Criminals per 100k")
```
No better relationship possible (sqrt etc i tried several), This lead us 
to the assumption that there is a saturation of the relationship after a 
certain point between criminals and beerhouses 

# Relationship of criminals and schools 
```{r scatterplots_schools_criminals}
ggplot(beer_data, aes(x = School, y = Criminals)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "loess", color = "red", span = 0.75) + 
  labs(title = "Crime Rate vs Schools (LOESS Curve)", 
       x = "Schools attendence per 10k", y = "Criminals per 100k")

```
linear relationship, something cosinus would maybe be appropriate but hard 
to calibrate therefore I leave it as it is. Often these transfomrations to 
not improve my model.

# log transfomration of the beerhouses 
```{r scatterplots_schools_criminals}
ggplot(beer_data, aes(x = ((Church)), y = Criminals)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "loess", color = "red", span = 0.75) + 
  labs(title = "Crime Rate vs Schools (LOESS Curve)", 
       x = "Schools attendence per 10k", y = "Criminals per 100k")

```
No linear relations ship at all, to me there is also no other relationship 
that I could identify 

```{r scatterplots}

```library(MASS)

# Fit a linear model for Box-Cox transformation
model <- lm(Criminals ~ BeerHouses, data = beer_data)

# Box-Cox transformation: Find optimal lambda
boxcox_result <- boxcox(model, lambda = seq(-0, 1.75, 0.1))
best_lambda <- boxcox_result$x[which.max(boxcox_result$y)]
print(paste("Best Lambda:", best_lambda))

# Apply the appropriate transformation
if (abs(best_lambda) < 0.01) {
  beer_data$School <- log(beer_data$School)  # Log transformation if λ ≈ 0
} else {
  beer_data$School <- (beer_data$School^best_lambda - 1) / best_lambda  # Box-Cox formula
}
```

#Best λ (Lambda) Value	Recommended Transformation
#λ = 1	No transformation needed (keep raw data).
#λ = 0	Log transformation: data$Criminals <- log(data$Criminals + 1)
#λ = 0.5	Square root transformation: data$Criminals <- sqrt(data$Criminals)
#λ = -1	Inverse transformation: data$Criminals <- 1 / data$Criminals
#λ = 2	Square transformation: data$Criminals <- data$Criminals^2

```{r scatterplots}

ggplot(beer_data, aes(x = 4^(1/(BeerHouses)), y = Criminals)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "loess", color = "red", span = 0.75) + 
  labs(title = "Crime Rate vs Beer Houses (LOESS Curve)", 
       x = "Beer Houses per 100k", y = "Criminals per 100k")

```

```{r scatterplots}
ggplot(beer_data, aes(x = School, y = Criminals)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "loess", color = "red", span = 0.75) + 
  labs(title = "Crime Rate vs Schools  (LOESS Curve)", 
       x = "Beer Houses per 100k", y = "Criminals per 100k")

```

```{r scatterplots}
ggplot(beer_data, aes(x = Church, y = Criminals)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "loess", color = "red", span = 0.75) + 
  labs(title = "Crime Rate vs Churchs (LOESS Curve)", 
       x = "Beer Houses per 100k", y = "Criminals per 100k")

```

# test for influence on criminality of every feature
# Linear models for each individual feature
```{r scatterplots}
model_beer <- lm(Criminals ~ BeerHouses, data = beer_data)
model_school <- lm(Criminals ~ School, data = beer_data)
model_church <- lm(Criminals ~ Church, data = beer_data)

# Summary of each model (shows t-values and p-values)
summary(model_beer)
summary(model_school)
summary(model_church)
```

Model: Criminals ~ BeerHouses
Coefficient (slope): 0.1162
t-value: 3.219
p-value: 0.00263 

H₀: β<sub>BeerHouses</sub> = 0 → Beer house density has no effect on crime rate
H₁: β<sub>BeerHouses</sub> ≠ 0 → Beer house density affects crime rate

Conclusion:
Since p = 0.00263 < 0.05, we reject H₀.
There is statistically significant evidence that beer house density is positively associated with crime rates.

Model: Criminals ~ School
Coefficient (slope): -0.05902
t-value: -1.457
p-value: 0.153 

H₀: β<sub>School</sub> = 0 → School attendance has no effect on crime rate
H₁: β<sub>School</sub> ≠ 0 → School attendance affects crime rate

Conclusion:
Since p = 0.153 > 0.05, we fail to reject H₀.
There is not enough evidence to conclude that school attendance is significantly associated with crime in this model.

Model: Criminals ~ Church
Coefficient (slope): 0.0008556
t-value: 0.022
p-value: 0.983

H₀: β<sub>Church</sub> = 0 → Church attendance has no effect on crime rate
H₁: β<sub>Church</sub> ≠ 0 → Church attendance affects crime rate

Conclusion:
Since p = 0.983 > 0.05, we fail to reject H₀.
There is no evidence of any relationship between church attendance and crime. The coefficient is near zero and clearly insignificant.

# Fit model 
```{r models}

lm1 <- lm(Criminals ~ BeerHouses + School, data=beer_data)  # Linear
lm2 <- lm(Criminals ~ log(BeerHouses) + School, data=beer_data)  # Log-transform BeerHouses
lm3 <- lm(Criminals ~ BeerHouses +  School , datare=beer_data)  # Quadratic
lm4 <- lm(Criminals ~ BeerHouses + I(BeerHouses^2) + School, datare=beer_data)  # Quadratic
lm5 <- lm(Criminals ~ BeerHouses + School + I(School^2), data=beer_data)  # Quadratic School
lm6 <- lm(Criminals ~ BeerHouses + cos(School), data=beer_data)  # Quadratic School
lm7 <- lm(Criminals ~ BeerHouses, data=beer_data)  # Linear
lm8 <- lm(Criminals ~ log(BeerHouses), data=beer_data)  # Linear
lm9 <- lm(Criminals ~ log(BeerHouses) + BeerHouses, data=beer_data)  # Linear

AIC(lm1, lm2, lm3, lm4, lm5, lm6, lm7, lm8,lm9)  # Lower AIC = better model
BIC(lm1, lm2, lm3, lm4, lm5, lm6,lm7, lm8,lm9)
```
    df      AIC
lm1  4 404.0409
lm2  4 400.2305
lm3  5 401.9763
lm4  5 405.3327
lm5  5 405.3327
lm6  4 408.7522
lm7  3 406.7524
lm8  3 403.5027
lm9  4 403.3548
    df      BIC
lm1  4 410.7964
lm2  4 406.9860
lm3  5 410.4207
lm4  5 413.7771
lm5  5 413.7771
lm6  4 415.5077
lm7  3 411.8190
lm8  3 408.5694
lm9  4 410.1103

r$>

even that schoolattendance does not ahve a big influence it reduces aic and bic a bit 
therefore model lm2 is best 


# Model Fitting
```{r residuals}
plot(lm2$fitted.values, lm2$residuals,
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

```
```{r residuals}

qqnorm(lm2$residuals)
qqline(lm2$residuals, col = "red")
```

```{r residuals}

plot(lm2, which = 4)  # Cook's distance
```
Points with Cook’s distance > 0.5 are usually considered moderately influential.

Points with Cook’s distance > 1 are highly influential (often candidates for investigation or exclusion).
Ther

e are some outliers to the right side of the plot 
this indicates some outliers.
Overall the model seems not to be perfect but a good fit. 

# Model Assessment
```{r residuals}

lm2_clean <- lm(Criminals ~ log(BeerHouses) + School, data = beer_data[-24, ])
summary(lm2_clean)
summary(lm2)
```

```{r residuals}
plot(lm2_clean$fitted.values, lm2_clean$residuals,
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

```
```{r residuals}

qqnorm(lm2_clean$residuals)
qqline(lm2_clean$residuals, col = "red")
```

Therefore the model y = log(beerhouses) + schoolattendance seems to be the best models
removing the moderate outliers reduces the normality of the data show in the two 
QQ-plots. 

## Final Model Representation

The linear regression model is fitted using least squares estimation.
For the model to be valid, the following assumptions must hold for the residuals:

Linearity

    The relationship between the predictors and the response is linear in the parameters.

    Assessed visually using scatterplots and residuals vs fitted plots.

Independence

    The residuals are independent from one another.

    Reasonable here as each county is treated as an independent observation.

Homoscedasticity

    The residuals have constant variance across all levels of fitted values.

    Assessed using the residuals vs fitted plot.

    In our model, no strong funnel shape is observed, so this assumption is reasonably met.

Normality of Residuals

    The residuals are normally distributed.

    Checked using Q-Q plots.

    There is some deviation in the tails, especially with the full dataset.
    However, after removing the influential observation (#24), the normality improves slightly.

# Conclusions

- **Beer house density is positively associated with crime**: More beer houses per capita correlate with higher crime rates.
- **School attendance is negatively associated with crime**: Higher education levels are linked to lower crime.
- **Church attendance was not statistically significant**: Once education and alcohol availability are considered, religion does not appear to predict crime rates.

This model explains approximately 30% of the variation in crime, suggesting other factors influence crime as well. The findings align with historical social theories that increasing education and regulating alcohol outlets may help reduce crime.

---

# References

- John Clay (1857). *On the Relation Between Crime, Popular Instruction, Attendance on Religious Worship, and Beer-House*. Journal of the Statistical Society of London, Vol. 20, pp. 22-32.
